P4IRS: A Code Generation Framework for Parallel and Performance-Portable Particle Simulations

### 2025-05-08 sage_latex_template/manuscript.tex
Improve added texts
Signed-off-by: Rafael Ravedutti Lucio Machado <rafael.r.ravedutti@fau.de>

\begin{figure*}[htb]
  \centering
  \includegraphics[width=0.75\textwidth]{pairs_architecture.pdf}
  \caption{\RMchange{P4IRS architecture. The user specifies the Particle Computations to be performed, which are compiled to the Python AST and further mapped to the P4IRS IR. Optimization Choices and Simulation Setup methods are used to determine which internal kernels are utilized, and the P4IRS IR is constructed in-place for the Internal Particle Optimization Kernels based on the inputs. Many Compilation Passes are performed to transform the P4IRS IR before triggering the Code Generator, which generates the Optimized Kernels, Data Structures and the PairsSimulation interface class. When whole-program generation is not used, a backend Simulation Code is required, and it uses the PairsSimulation interface to call P4IRS generated kernels. The final Binary is then generated with any backend compiler the user chooses (i.e., GCC, CLANG, ...), and it must be linked with the Runtime Library. The Runtime Library uses the P4IRS Metadata structure to handle different particle data structures and adapt to different simulations.}}
  \label{fig:pairs_arch}
\end{figure*}

@@

\RMchange{This and the following sections focus on describing the architecture from the P4IRS compiler.
\autoref{fig:pairs_arch} depicts a general overview of the P4IRS architecture from when the user inputs are given until the final binary is generated.

An important aspect of a \ac{DSL} is to be able to use domain knowledge for their compiler transformations, thus enabling code optimizations that may not be detected by general purpose compilers.
In order to achieve this, P4IRS provides its own \ac{IR} which can represent elements that are specific to particle computations.
Note that we use the term \ac{IR} in this paper to denominate the internal data structure utilized to represent the parsed code.
Our \ac{IR} is not formally defined as is usually the case for most \ac{IR}s from general purpose compilers, and it is not our intention here to provide a formal and strong definition for it.
Instead, we present an informal and rather simple \ac{AST} representation which contains elements associated with particle computations and data structures.
These particle-specific elements in the AST can then be used in P4IRS compilation passes, and at some point during the compilation are lowered, resulting in the low level AST that corresponds to the code implementation that matches the strategies given by the user.}

@@

\begin{figure*}[htb]
  \centering
  \includegraphics[width=0.75\textwidth]{pairs_ir.pdf}
  \caption{\RMchange{P4IRS IR for the BuildNeighborLists example. Red boxes represent elements that are specific to particle simulations, and are used for domain-specific transformations in the \ac{AST}, at some point these elements are lowered to generic, primitive elements that should be straightforward to map into code in a general-purpose language or even an IR such as the LLVM IR. Green boxes represent some of these lower level, generic elements. Blue boxes represent symbols from the simulation, and the dashed lines in red show when these symbols come from a particle-specific element. Yellow boxes represent constants, and both ``s'' elements represent a pattern that is repeated in the constructed \ac{AST} according to the referenced loop in their respective orange boxes.}}
  \label{fig:pairs_ir}
\end{figure*}

\RMchange{The P4IRS IR can be constructed in two ways: (a) by a mapping from the Python IR generated through the \emph{compile()} method on Python functions and (b) by explicitly constructing the IR elements in-place.
The former is used for user-defined particle interactions (like the \emph{lennard\_jones} method from \autoref{lst:pairslj}) and the latter is normally used internally by P4IRS to construct the IR for internal routines.}
\autoref{lst:build_neigh} displays an example for constructing the intermediate representation that generates the code that builds the neighbor-lists for each particle in the system.
The \emph{BuildNeighborLists} class is a separate, high-level element from the AST in P4IRS.
During the compilation process, P4IRS performs analyses and transformations to lower this element by using its \emph{lower} method, and decorators such as the \emph{@pairs\_device\_block} are used to determine the target hardware where the code must be executed.

\autoref{fig:pairs_ir} depicts a schematic diagram of the constructed P4IRS \ac{IR} for the \emph{BuildNeighborLists} example.
\RMchange{Note the domain-specific elements in the IR such as \emph{ParticleInteraction} and \emph{ParticleFor}, as already discussed these can be used during compilation passes for obtaining domain knowledge, and at some point they are lowered according to the strategies specified by the user.
For instance, the \emph{ParticleInteraction} abstract element can be lowered to a code representation that traverses pair of particles by either using the Linked Cells or Verlet Lists data structures.}
%Other elements such as the \emph{ParticleFor} and \emph{ParticleInteraction} are used as higher-level elements of the AST.
%These will be lowered at some point based on their strategies and given parameters.
In this case, the \emph{ParticleInteraction} is lowered by using the linked-cells data structure to build the neighbor-lists as is done in traditional implementations.
The interaction object also contains methods to access the current particle (i), the neighbor particle (j) and the shape for the neighbor particle (shape).
The latter is particulary important for the shape-partitioning and multi-kernel code generation discussed in the \nameref{sec:shape_partitioning} section.

@@

\RMchange{
The final step of the P4IRS compiler is the code generation, which converts the optimized \ac{IR} to low-level hardware assembly instructions or to another high-level language (source-to-source compilation).
In its current state, P4IRS generates C++ and CUDA code with MPI for distributed-memory parallelism, but new backend languages, intermediate representations and technologies can be included without difficulties by extending the code generator.
Although most of the code generation is straightforward for the \ac{IR} in its lowest level, there are some important points that are important to handle different backends and achieve performance-portability.

@@

\RMchange{The runtime library must be linked with the generated binary during its execution, and therefore it must be properly built for the target hardware and platform, and either installed in the target system or its path included in the list of paths where dynamic libraries are looked for (for example, the \emph{LD\_LIBRARY\_PATH} environment variable on Linux systems).
Although its compilation is done separately from the generated code and it should ideally be dynamically linked during runtime, it is possible to compile both together and statically link the runtime library code with the simulation code, hence generating a single, platform-specific, stand-alone binary that does not require further linking with the runtime code.}

@@

\begin{table*}[htb]
    \centering
    \begin{tabular}{c|c|c|c|c}
%        \hline
        Variant & pairs-fn & mdbench-fn & pairs-hn & mdbench-hn \\
        \hline
        Ice Lake & 0.962 & 1.019 & 1.662 & 1.595 \\
        Milan & 0.908 & 0.883 & 1.354 & 1.367 \\
        A40  & 26.452 & 27.025 & 29.688 & --- \\
        A100 & 75.113 & 87.381 & 87.968 & --- \\
%        \hline
    \end{tabular}
    \caption{\RMchange{Summary of performance results (in million particle updates per second) for each P4IRS and MD-Bench simulation runs of the MD benchmark on a single core of Ice Lake and Milan CPUs, and one A40 and A100 GPUs.}}
    \label{tab:md_particle_updates_per_second_md}
\end{table*}

\RMchange{\autoref{tab:md_particle_updates_per_second_md} summarizes the results in million particle updates per second running on a single CPU core and on a single GPU for the \ac{MD} case.
Note that our goal is not to compare the performance on a single CPU core against the performance on a single GPU since this is not a fair comparison.
However, we want to demonstrate that we can achieve a significant performance improvement when running P4IRS code on a GPU, and the results are close (and better in some cases) when comparing P4IRS code to MD-Bench on all the systems listed.}

@@

\begin{table*}[htb]
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
        Variant & pairs-fn & pairs-fn-nb & pairs-fn-2r & pairs-fn-nb-2r & pairs-fn-mk & pairs-fn-nb-mk & mesapd \\
        \hline
        Ice Lake & 76.573 & 48.549 & 164.202 & 101.062 & --- & --- & 47.304 \\
        Milan & 113.229 & 68.008 & 225.374 & 134.969 & --- & --- & 64.347 \\
        A40 & 68.866 & 83.06 & --- & --- & 63.835 & 79.828 & --- \\
        A100 & 210.41 & 235.753 & --- & --- & 209.969 & 251.292 & --- \\
    \end{tabular}
    \caption{\RMchange{Summary of performance results (in million particle updates per second) for each P4IRS and MESA-PD simulation runs of the DEM benchmark on a single core of Ice Lake and Milan CPUs, and one A40 and A100 GPUs.}}
    \label{tab:dem_particle_updates_per_second_dem}
\end{table*}

\RMchange{\autoref{tab:dem_particle_updates_per_second_dem} summarizes the results running on a single and two CPU cores, as well as on a single GPU, the results are shown in million particle updates per second.
The smaller number of overall particle interactions combined with a more complex kernel makes the Linear Spring-Dashpot method less suitable for GPUs when using the standard Linked Cells strategy.
Better performance is obtained when both the Neighbor-Lists per Cell and the Multi-Kernel approach are used for GPUs, but there is still room for improvement.
Evaluating and designing new algorithms for calculating this specific kernel efficiently on GPUs is one of our next goals, and code generation can be particularly helpful because we can do experimentation with a single high-level description of the problem.}

>>>>

\begin{figure*}[htb]
  \centering
  \includegraphics[width=0.75\textwidth]{pairs_architecture.pdf}
  \caption{\RMchange{P4IRS architecture. The user specifies the Particle Computations to be performed, which are compiled to the Python AST and further mapped to the P4IRS IR. Optimization Choices and Simulation Setup methods are used to determine which internal kernels are utilized, and the P4IRS IR is constructed in-place for the Internal Particle Optimization Kernels based on the inputs. Many Compilation Passes are performed to transform the P4IRS IR before triggering the Code Generator, which generates the Optimized Kernels, Data Structures and the PairsSimulation interface class. If whole-program generation is not used, a backend Simulation Code is required, which uses the PairsSimulation interface to call P4IRS generated kernels. The final Binary is then generated with any backend compiler of the user's choice (i.e., GCC, CLANG, ...), and it must be linked with the Runtime Library. The Runtime Library uses the P4IRS Metadata structure to handle different particle data structures and to adapt to different simulations.}}
  \label{fig:pairs_arch}
\end{figure*}

@@

\RMchange{This and the following sections focus on describing the architecture from the P4IRS compiler.
\autoref{fig:pairs_arch} depicts a general overview of the P4IRS architecture from the time the user input is given to the time the final binary is generated.

An important aspect of a \ac{DSL} is to be able to use domain knowledge for its compiler transformations, thus enabling code optimizations that may not be detected by general purpose compilers.
In order to achieve this, P4IRS provides its own \ac{IR} which can represent elements that are specific to particle computations.
Note that in this paper we use the term \ac{IR} to denominate the internal data structure utilized to represent the parsed code.
Our \ac{IR} is not formally defined, as is usually the case for most \ac{IR}s from general-purpose compilers, and it is not our intention here to provide a formal and rigorous definition for it.
Instead, we present an informal and rather simple \ac{AST} representation that provide elements associated with particle computations and data structures.
These particle-specific elements in the AST can then be used in P4IRS compilation passes, and at some point during the compilation are lowered, resulting in the low-level AST that corresponds to the code implementation that matches the strategies given by the user.}

@@

\begin{figure*}[htb]
  \centering
  \includegraphics[width=0.75\textwidth]{pairs_ir.pdf}
  \caption{\RMchange{P4IRS IR for the BuildNeighborLists example. Red boxes represent elements that are specific to particle simulations, and are used for domain-specific transformations in the \ac{AST}, at some point these elements are lowered to generic, primitive elements that should be straightforward to map to code in a general-purpose language or even an IR such as the LLVM IR. Green boxes represent some of these lower level, generic elements. Blue boxes represent symbols from the simulation, and the dashed red lines show when these symbols come from a particle-specific element. Yellow boxes represent constants, and the two ``s'' elements represent patterns that are repeated in the constructed \ac{AST} according to the referenced loop in their respective orange boxes.}}
  \label{fig:pairs_ir}
\end{figure*}

\RMchange{The P4IRS IR can be constructed in two ways: (a) by mapping the Python IR generated with the \emph{compile()} method on Python functions, and (b) by explicitly constructing the IR elements in-place.
The former is used for custom particle interactions (such as the \emph{lennard\_jones} method from \autoref{lst:pairslj}) and the latter is normally used internally by P4IRS to construct the IR for internal routines.}
\autoref{lst:build_neigh} displays an example for constructing the intermediate representation that generates the code that builds the neighbor-lists for each particle in the system.
The \emph{BuildNeighborLists} class is a separate, high-level element from the AST in P4IRS.
During the compilation process, P4IRS performs analyses and transformations to lower this element by using its \emph{lower} method, and decorators such as the \emph{@pairs\_device\_block} are used to determine the target hardware where the code must be executed.

\autoref{fig:pairs_ir} depicts a schematic diagram of the constructed P4IRS \ac{IR} for the \emph{BuildNeighborLists} example.
\RMchange{Note the domain-specific elements in the IR, such as \emph{ParticleInteraction} and \emph{ParticleFor}; as discussed earlier, these can be used during compilation passes to gain domain knowledge, and at some point they are lowered according to the user-specified strategies.
For instance, the \emph{ParticleInteraction} abstract element can be lowered to a code representation that traverses pairs of particles using either the Linked Cells or Verlet Lists data structures.}
%Other elements such as the \emph{ParticleFor} and \emph{ParticleInteraction} are used as higher-level elements of the AST.
%These will be lowered at some point based on their strategies and given parameters.
In this case, the \emph{ParticleInteraction} is lowered by using the linked-cells data structure to build the neighbor-lists as is done in traditional implementations.
The interaction object also contains methods to access the current particle (i), the neighbor particle (j) and the shape for the neighbor particle (shape).
The latter is particulary important for the shape-partitioning and multi-kernel code generation discussed in the \nameref{sec:shape_partitioning} section.

@@

\RMchange{
The final step of the P4IRS compiler is the code generation, which converts the optimized \ac{IR} into low-level hardware assembly instructions or to another high-level language (source-to-source compilation).
In its current state, P4IRS generates C++ and CUDA code with MPI for distributed memory parallelism, but new backend languages, intermediate representations and technologies can be included without difficulties by extending the code generator.
Although most of the code generation is straightforward for the \ac{IR} at its lowest level, there are important aspects that need to be considered to handle different backends and achieve performance portability.

@@

\RMchange{The runtime library must be linked to the generated binary during execution, so it must be properly built for the target hardware and platform, and either installed on the target system or its path must be included in the list of paths where dynamic libraries are searched for (for example, the \emph{LD\_LIBRARY\_PATH} environment variable on Linux systems).
Although it is compiled separately from the generated code and should ideally be linked dynamically at runtime, it is possible to compile both together and statically link the runtime library code with the simulation code, thus generating a stand-alone, platform-specific binary that does not require further linking with the runtime code.}

@@

\begin{table*}[htb]
    \centering
    \begin{tabular}{c|c|c|c|c}
%        \hline
        Variant & pairs-fn & mdbench-fn & pairs-hn & mdbench-hn \\
        \hline
        Ice Lake & 0.962 & 1.019 & 1.662 & 1.595 \\
        Milan & 0.908 & 0.883 & 1.354 & 1.367 \\
        A40  & 26.452 & 27.025 & 29.688 & --- \\
        A100 & 75.113 & 87.381 & 87.968 & --- \\
%        \hline
    \end{tabular}
    \caption{\RMchange{Summary of performance results (in millions of particle updates per second) for each P4IRS and MD-Bench simulation run of the MD benchmark on a single core of the Ice Lake CPU, a single core of the Milan CPU, a single A40 GPU, and a single A100 GPU.}}
    \label{tab:md_particle_updates_per_second_md}
\end{table*}

\RMchange{\autoref{tab:md_particle_updates_per_second_md} summarizes the results in million particle updates per second, running on a single CPU core and on a single GPU for the \ac{MD} case.
Note that our goal is not to compare performance on a single CPU core to performance on a single GPU, as this is not a fair comparison.
However, we do want to show that we can achieve a significant performance improvement when running P4IRS code on a GPU, and the results are close (and in some cases better) when comparing P4IRS code to MD-Bench on all the systems listed.}

@@

\begin{table*}[htb]
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|c}
        Variant & pairs-fn & pairs-fn-nb & pairs-fn-2r & pairs-fn-nb-2r & pairs-fn-mk & pairs-fn-nb-mk & mesapd \\
        \hline
        Ice Lake & 76.573 & 48.549 & 164.202 & 101.062 & --- & --- & 47.304 \\
        Milan & 113.229 & 68.008 & 225.374 & 134.969 & --- & --- & 64.347 \\
        A40 & 68.866 & 83.06 & --- & --- & 63.835 & 79.828 & --- \\
        A100 & 210.41 & 235.753 & --- & --- & 209.969 & 251.292 & --- \\
    \end{tabular}
    \caption{\RMchange{Summary of performance results (in millions of particle updates per second) for each P4IRS and MESA-PD simulation run of the DEM benchmark on a single core of Ice Lake CPU, on a single core of a Milan CPU, on a single A40 GPU, and on a single A100 GPU.}}
    \label{tab:dem_particle_updates_per_second_dem}
\end{table*}

\RMchange{\autoref{tab:dem_particle_updates_per_second_dem} summarizes the results running on one and two CPU cores, as well as on a single GPU, the results are shown in millions of particle updates per second.
The lower total number of overall particle interactions combined with a more complex kernel makes the Linear Spring-Dashpot method less suitable for GPUs when using the standard Linked Cells strategy.
Better performance is achieved using both the Neighbor-Lists per Cell and the Multi-Kernel approaches for GPUs, but there is still room for improvement.
Evaluating and designing new algorithms to efficiently compute this specific kernel on GPUs is one of our next goals, and code generation can be particularly helpful because we can experiment with a single high-level description of the problem.}
